$\pi$ with leftovers: \\ a mechanisation in Agda

### 2020-05-01 paper/paper.tex

Structural congruence is a congruent equivalence relation on processes.
\autoref{fig:struct-cong-base} lists the base cases of structural congruence.
The type $\Unused_i \; Q$ is an inductive proof that witnesses that the variable index $i$ does not appear nor in the inputs nor in the outputs of $Q$.
The function $lower_i \; Q \; uQ$ traverses $Q$ decrementing every index bigger than $i$.
Finally, $swap_i \; P$ traverses $P$ (of type $\Process_{\suc \suc n}$) and swaps variable references $i$ and $\suc i$.
In all three, $i$ is incremented every time we go under a binder.

\begin{figure}[h]
\begin{mathpar}
  \datatype
  { }
  {P \eqeq Q : et}
  \; \textsc{StructCong}

@@

\begin{nilemma}
  \label{lm:lower-lift}
  The function $lower_i \; P \; uP$ has an inverse $lift_i \; P$ that increments every $\textsc{Var}$ greater than or equal to $i$, such that $lift_i \; (lower_i \; P \; uP) \equiv P$.
\end{nilemma}
\begin{proof}
  By structural induction on \textsc{Process} and \textsc{Var}.
\end{proof}

\begin{nilemma}
  \label{lm:swap-swap}
  The function $swap_i \; P$ is its own inverse: $swap_i \; (swap_i \; P) \equiv P$.
\end{nilemma}
\begin{proof}
  By structural induction on \textsc{Process} and \textsc{Var}.
\end{proof}

Structural congruence is a congruent equivalence relation.
As such, rewrites can happen anywhere inside a process, and are closed under reflexivity, symmetry and transitivity as shown in \autoref{fig:struct-cong}.
In  \autoref{subject-congruence} we prove that if two processes $P$ and $Q$ are structurally congruent and $P$ is well-typed, then so is $Q$.
In the transitivity case, we must show that if $P$ is structurally congruent to $Q$ and $Q$ is to $R$, and $P$ is well-typed, then so is $R$.
To do so, we need to proceed by induction and first get a proof of the well-typedness of $Q$, then use that to reach $R$.
To show the typechecker that the doubly recursive call terminates we index the equivalence relation $=$ by a type $\Rec$ that models the structure of the recursion.

\begin{figure}[h]
\begin{mathpar}
  \datatype
  { }
  {\Rec : et}
  \; \textsc{Rec}

  \inferrule
  { }
  {\constr{zero} : \Rec}

  \inferrule
  {r : \Rec}
  {\constr{one} \; r : \Rec}

  \inferrule
  {r \; s : \Rec}
  {\constr{two} \; r \; s : \Rec}

  \datatype
  {P \, Q : \Process_n \\ r : \Rec}
  {P \eq{r} Q : et}
  \; \textsc{Equals}

  \inferrule
  {eq : P \eqeq Q}
  {\constr{struct} \; eq : P \eq{\constr{zero}} Q}

>>>>

The complete structural congruence is given in \autoref{app:struct}.
Figure \ref{fig:struct-cong-base} lists the base cases.

\todo{remove space between inferrules}
\begin{figure}[h]
\begin{mathpar}
  \datatype
  { }
  {P \eqeq Q : et}
  \; \textsc{StructCong}

@@

In $\constr{scope-ext}$ the type $\Unused_i \; Q$ is an inductive proof that witnesses that the variable index $i$ does not appear neither in the inputs nor in the outputs of $Q$.
The function $lower_i \; Q \; uQ$ traverses $Q$ decrementing every index bigger than $i$.
In $\constr{scope-comm}$ function $swap_i \; P$ traverses $P$ (of type $\Process_{\suc \suc n}$) and swaps variable references $i$ and $\suc i$.
In all the above, $i$ is incremented every time we go under a binder.

Structural congruence is a congruent equivalence relation.
As such, rewrites can happen anywhere inside a process, and are closed under reflexivity, symmetry and transitivity as shown by the first row of Figure \ref{fig:struct-cong}.
The rest of the rules defines structural congruence under a context $\mathcal{C}[\cdot]$ \cite{Sangio01}, respectively restriction, composition, input and output.
%In \S \ref{subject-congruence} we prove that if two processes $P$ and $Q$ are structurally congruent and $P$ is well-typed, then so is $Q$.

\begin{figure}[h]
\begin{mathpar}
  \inferrule
  { }
  {\constr{refl} : P \eq{\constr{zero}} P}

  \inferrule
  {eq_1 : P \eq{r} Q \\ \; eq_2 : Q \eq{s} R}
  {\constr{trans} \; eq_1 \; eq_2 : P \eq{\constr{two} \; r \; s} R}
